### 2020-09-03 thesis/thesis.tex
[thesis] Corrections in first two chapters

In this thesis we discuss issues arising from combining algebraic effects
and handlers with resource management, and present a solution through coalgebraic
means along with presentation of Freak, experimental programming language with
(co)algebraic effects and (co)handlers. Implementation is based on Continuation
Passing Style translation.

@@

They give the developer necessary tools for declarative programming with
computational effects on a high level of abstraction, by defining
effects as an API\@ in the code. At the same time, algebraic effects preserve
a lot of flexibility and have a strong theoretical foundations.

@@

In the above code we open file, and based on nondeterministic choice, we write
to the file, and then close it. This piece of code looks harmless, however, as
we invoke resumption for the second time, we are attempting to write to the
closed file, and then close it once again. In fact, dropping resumption is also
considered harmful, as we would not attempt to close the file.

@@

Under these circumstances one may pose a question, why do we still bother with
development of languages theory, since so much has been done already. Is there
anything that drives us towards further research? Indeed, one active branch
revolves around equational theory to assess equality of two programs, which we
know that in the general setting is undecidable. Proof methods may include
extensional, contextual or logical equality. However, there is no doubt that
these formal ways of reasoning about programs, while being crucial for assessing
correctness, do not bring direct benefits for everyday use cases, as they are
rarely accessible by a common developer.

Other branch of languages theory, that we shall investigate in this thesis,
is about taming complexity of programs. Various methods of static analysis has
been developed, most notably, type systems. Thanks to strong
and static type systems along with their implementations, we have solid tools to
work efficiently on functions that are pure. That being said, we claim that the
core complexity of programs comes from side effects, or more generally, computational
effects, which we cannot avoid in writing anything useful.

We need to have a good way for handling computational effects. One of the ways
to model them, can be done through monads~\cite{monads-wadler, moggi}. However,
they were found to be, to say the least, cumbersome to work with when the number
of different effects increase. It is perhaps not a coincidence that many
functional programming languages do not have them, because of the non-composable
nature of them, or at least not composable in their implementations.

\noindent
Let's put the following functions

@@

\noindent
where $m$, $m'$ are monads. Functions $f$, $g$ can be composed using standard
composition, for $f'$, $g'$ we can use Kleisli composition operator from monad $m$.
What in case of adding one more effect, $f''$ and $g''$?

It turns out, that it becomes complex and unpleasant to combine two functors
together to form a new monad, and for this purpose, monad
transformers~\cite{monad-transformers} arose in Haskell. Most of the common
languages avoid this by not expressing computational effects in the type system,
and instead one may think about functions as being implicitly embed in a Kleisli
Category over a functor $T$~\cite{kleisli}, where $T$ is a hidden signature
over all possible side effects that occur in our program.

@@

Algebraic effects can be thought of as an public interface for computational effects.
Declarative approach allow us to write programs in which the actual semantic of
source code is dependent on handler that defines the meaning of a subset of effects.

@@

Patterns like these are well known in programming, for which other alternatives
arose. One can mention interfaces from object-oriented programming, which are also
a way to describe the API of a certain component. In order to abstract from
implementation details, we pass an object around which represents a certain
interface. While this is certainly better than having no abstractions at all,
we need to pass this interface into every function that is going to use it.
This practice is called dependency injection, and while it sounds like it solves
some of the issues, we end up in functions that need to carry representants
of the interfaces, and pass them in every subcall that they make.

That being said, it's only one particular issue that algebraic effects address.
There are many other great sources for getting familiarized with effects from
practical side~\cite{handlers-tutorial, programming-in-eff, koka-tutorial},
so we will omit further explanations. More examples can be found in
Chapter~\ref{chapter:co-effectful-programming}.

@@

Algebraic effects can be described via operational means, however,
for the purpose of presenting the duality between algebra and coalgebra,
we allow ourselves to wander a bit deeper into category theory and describe
effects from denotational point of view~\cite{bauer-what-is-algebraic}.

@@

    $\mathVar{Free}$ $\mathVar{F-algebra}$ on an object $A$ (of generators) in $\mathcal{C}$ is meant an algebra

@@

    $ \varphi_{A} : F A^{\#} \longrightarrow A $

@@

    \noindent
    together with an universal arrow $ \eta_{A} : A \longrightarrow A^{\#} $.
    Universality means that for every algebra $ \beta : F B \longrightarrow B $
    and every morphism $ f : A \longrightarrow B $ in $ \mathcal{C}$, there
    exists a unique homomorphism $ \overline{f} : A^{\#} \longrightarrow B $
    extending $f$, i.e.\ a unique morphism in $ \mathcal{C}$ for which the diagram
    below commutes:

@@

    % == begin Sigma-terms ==
    \begin{definition}
    Collection of $\mathVar{\Sigma-terms}$ is a free algebra with a generator $X$
    for a functor $ \mu H_{\Sigma} $ that maps objects into trees over a given
    signature $ \Sigma $ and morphisms into folds over trees.

@@

    \noindent
    Interpretation may be naturally extended to $\mathVar{\Sigma-terms}$, such
    that a given $\mathVar{\Sigma-term}$ $X \mid t$ is interpreted by a map
    which sends variables into projections from environment and terms into
    map composition over each subterm.

@@

    Let $L$, $M$ be models of a theory $T$. $ \mathVar{T-homomorphism}$
    $\varphi : L \rightarrow M$ is a map, such that, for every operation symbol
    $op_{i}$ in $T$ the following diagram commutes:

@@

    \begin{definition} \textit{Free model} is just a model that is free algebra.
    \end{definition}

@@

>>>>

In this thesis we discuss issues arising from combining algebraic effects
and handlers with resource management, and present a solution through coalgebraic
means along with presentation of Freak, an experimental programming language with
(co)algebraic effects and (co)handlers. Implementation is based on Continuation
Passing Style translation.

@@

They give the developer necessary tools for declarative programming with
computational effects on a high level of abstraction, by defining
effects as an API\@ in the code. At the same time, algebraic effects preserve
a lot of flexibility and have strong theoretical foundations.

@@

In the code above we open a file, and based on nondeterministic choice, we write
to the file, and then close it. This piece of code looks harmless, however, as
we invoke resumption for the second time, we are attempting to write to the
closed file, and then close it once again. In fact, dropping resumption is also
considered harmful, as we would not attempt to close the file.

@@

Under these circumstances one may pose a question, why we still bother with
development of the theory of programming languages, since so much has been
done already. Is there
anything that drives us towards further research? Indeed, one active branch
revolves around equational reasoning to establish equality of two programs, which we
know is undecidable in general. Proof methods may include
extensional, contextual or logical equality. However, there is no doubt that
these formal ways of reasoning about programs, while being crucial for assessing
correctness, do not bring direct benefits for everyday use cases, as they are
rarely accessible by a common developer.

Other branch of languages theory that we shall investigate in this thesis
is about taming complexity of programs. Various methods of static analysis have
been developed, most notably, type systems. Thanks to strong
and static type systems along with their implementations, we have solid tools to
work efficiently on functions that are pure. That being said, we claim that the
core complexity of programs comes from side effects, or more generally, computational
effects, which we cannot avoid in writing anything useful.

We need to have a good way for handling computational effects. One of the ways
to model them can be done through monads~\cite{monads-wadler, moggi}. However,
they were found to be, to say the least, cumbersome to work with when the number
of different effects increase. It is perhaps not a coincidence that many
functional programming languages do not have them, because of their non-composable
nature.

\noindent
For example, consider the following functions

@@

\noindent
where $m$, $m'$ are monads. Functions $f$, $g$ can be composed using standard
composition, for $f'$, $g'$ we can use Kleisli composition operator from monad $m$.
How about the case of adding one more effect, $f''$ and $g''$?

It turns out that it may become complex and unpleasant to combine two monads
together to form a new monad, and for this purpose, monad
transformers~\cite{monad-transformers} arose in Haskell. Most of the usual
languages avoid this by not expressing computational effects in the type system,
and instead one may think about functions as being implicitly embedded in a Kleisli
Category of a functor $T$~\cite{kleisli}, where $T$ is a hidden signature
of all possible side effects that may occur in our program.

@@

Algebraic effects can be thought of as a public interface for computational effects.
Declarative approach allows us to write programs in which the actual semantics of
source code depends on a handler that defines the meaning of a subset of effects.

@@

Patterns like these are well known in programming, for which other alternatives
arose. One can mention interfaces from object-oriented programming, which are also
a way to describe the API of a certain component. In order to abstract from
implementation details, we pass an object around which represents a certain
interface. While this is certainly better than having no abstractions at all,
we need to pass this interface to every function that is going to use it.
This practice is called dependency injection, and while it sounds like it solves
some of the issues, we end up in functions that need to carry representants
of the interfaces, and pass them in every subcall that they make.

That being said, it's only one particular issue that algebraic effects address.
There are many other great sources for getting familiarized with effects from the
practical side~\cite{handlers-tutorial, programming-in-eff, koka-tutorial},
so we will omit further explanations. More examples can be found in
Chapter~\ref{chapter:co-effectful-programming}.

@@

Algebraic effects can be described via operational means, however,
for the purpose of presenting the duality between algebra and coalgebra,
we allow ourselves to wander a bit deeper into category theory and describe
effects from the denotational point of view~\cite{bauer-what-is-algebraic}.

@@

    For an endofunctor $F$, a $\mathVar{Free}$ $\mathVar{F-algebra}$ on an
    object $A$ (of generators) in $\mathcal{C}$ is an algebra

@@

    $ \varphi_{A} : F A^{\#} \longrightarrow A^{\#} $

@@

    \noindent
    together with a universal arrow $ \eta_{A} : A \longrightarrow A^{\#} $.
    Universality means that for every algebra $ \beta : F B \longrightarrow B $
    and every morphism $ f : A \longrightarrow B $ in $ \mathcal{C}$, there
    exists a unique homomorphism $ \overline{f} : A^{\#} \longrightarrow B $
    extending $f$, i.e.\ a unique morphism in $ \mathcal{C}$ for which the diagram
    below commutes:

@@

    % == begin Sigma-terms ==
    \begin{definition}
    A collection of $\mathVar{\Sigma-terms}$ is a free algebra with a generator $X$
    for a functor $ \mu H_{\Sigma} $ that maps objects into trees over a given
    signature $ \Sigma $ and morphisms into folds over trees.

@@

    \noindent
    Interpretation may be naturally extended to $\mathVar{\Sigma-terms}$, such
    that a given $\mathVar{\Sigma-term}$ $X \mid t$ is interpreted by a map
    which sends variables to projections from environment and terms to
    map composition over each subterm.

@@

    Let $L$, $M$ be models of a theory $T$. A $\mathVar{T-homomorphism}$
    $\varphi : L \rightarrow M$ is a map, such that for every operation symbol
    $op_{i}$ in $T$, the following diagram commutes:

@@

    \begin{definition} A \textit{Free model} is just a model that is a free algebra.
    \end{definition}

@@

% \vbox{\begin{verbatim}
%     handle
%         let fh <- observe Open "praise.txt" in
%         let c <- do Choice () in
%         let text <- if c then return "Guy Fieri" else return "is cool" in
%         let _ <- observe Write (fh, text) in
%         observe Close fh
%     with {
%         Choose _ r ->
%             let t <- r 1 in
%             let f <- r 0 in
%             return (t, f) |
%         return x -> return x
%     }
% \end{verbatim}
% }
